---
title: "Evolutionary rescue in flooding events"
author: "Ramin Dürst & Matthias Schönholzer"
date: "16.10.2024"
output:
  html_document:
    df_print: paged
co-author: "Base code by Claudia Bank"
---

## List of variables and parameters

-   $N_a$ current size of the wild-type (a) population (changes over time)
-   $N_A$ current size of the mutant (A) population (changes over time)
-   `N_init_a` initial population size of wild-type population (should be a positive integer)
-   `N_init_A` initial population size of the mutant population (~~should be a non-negative integer~~ –\> will be 0 at start)
-   `decay_rate` detrimental effect of the environment on the wild type (i.e., the wild-type population decays at this rate; should be \>0 to induce potential extinction)
-   `sel_coeff` ~~selective advantage of the mutant over the wild type (should be \>decay_rate to allow for evolutionary rescue)~~ –\> selection coefficient of wild type declines during the event, while the selection coefficient of the mutant increases during an event.
-   `mut_rate` mutation rate from the wild type to the mutant genotype (should be \>0 and small)
-   `t_max` number of generations after which to stop the simulation (should be an integer \>0; this is not a model parameter but a setting for the simulations). In the code presented here, the simulation stops automatically if the population goes extinct, or if it recovers and exceeds 1.5 times the original total population size, so `t_max` can be set to a large value without creating extremely long simulation times.
-   `duration of events` the timespan of generations the event occurs. Depends on the decay rate of the wild type and will be over 1.
-   `frequency of events` at which frequency the events occur over time
-   `strength of events` how strong the events affect the population. Is a combination of decay / growth rate and selection coefficient.

## Scenario

A hypothetical population of plants in a water meadow by a river with two genotypes: one which does bad during flooding (wild type) and one which does good during flooding (mutant). The frequency of being flooded increases due to climate change and depending on frequency, either the wild type persists, the mutant takes over, or both genotypes and therefor the population go extinct.

## Simulate one generation of population size change

We want to observe the numbers of wild-type and mutant individuals over time. As first step, we simulate one generation of the population. At time $t$, the population has $N_a$ wild-type individuals and $N_A$ mutant individuals. First, these reproduce. This is modeled by drawing a Poisson random variable according to the expected absolute number $N_a \cdot (1-r)$ of wild-type individuals, where $r$ is the decay rate of the wild type, and $N_A \cdot (1-r+s)$ of mutant individuals, where $s$ is the selective advantage of the mutant over the wild time. Then mutation from $a$ to $A$ occurs as a Poisson random variable according to the given mutation rate. This defines the composition of the population at time $t+1$.

```{r}
#Original

simulate_one_gen <- function(N_a, N_A, decay_rate, sel_coeff, mut_rate) {
  # draw offspring according to Poisson distribution
  offsp_a <- rpois(1, N_a * (1-decay_rate))
  offsp_A <- rpois(1, N_A * (1-decay_rate+sel_coeff))
  # draw new mutants according to Poisson distribution
  mut_a_to_A <- rpois(1, offsp_a * mut_rate)
  
  # determine new population sizes of wild type and mutant
  N_a_new <- max(offsp_a - mut_a_to_A, 0)
  N_A_new <-  offsp_A + mut_a_to_A
  
  return(c(N_a_new, N_A_new))
}
# Test the function
print(simulate_one_gen(100,0,0.1,0.2,0.0))
```

## Simulate a population trajectory

We can now track the dynamics of the population from one generation to another, but this is not enough. We want to follow the population from time 0 until it either is rescued or goes extinct. This is what the following function does.

```{r}
#Original

simulate_pop <- function(N_init_a, N_init_A, decay_rate, sel_coeff, mut_rate, t_max) {
  # Create the vector in which to save the results
  pop_vector <- c(N_init_a, N_init_A)
  # initiate the variables
  pop_new <- c(N_init_a, N_init_A)
  
  # run the simulation until generation t_max
  for (i in 1:t_max+1) {
    # redefine the current population one generation later
    pop_new <- simulate_one_gen(pop_new[1],pop_new[2], decay_rate, sel_coeff, mut_rate)
    # add the new population sizes to the output vector
    pop_vector <- rbind(pop_vector,pop_new)
    # condition to stop the simulation before t_max: either the population exceeds 1.5 times the original population size, or it goes extinct
    if (pop_new[1]+pop_new[2]>=1.5*(N_init_a+ N_init_A) | pop_new[1]+pop_new[2]==0) break
  }
  
  # define the row and column names of the output vector
  rownames(pop_vector) <- (0:t_max)[1:length(pop_vector[,1])] # note that the vector has to be cut if the simulation stopped early
  colnames(pop_vector) <- c("a","A")
  # return the result
  return(pop_vector)	
}


# Test the function and plot the result
# set t_max to a value to be used througout
max_gen <- 1000
# create your simulation data, without mut_rate & sel_coeff and a higher decay_rate
output <- simulate_pop(1000,0,0.1,0.2,0.001,max_gen)
# show the last few lines of the data table
print(output)
# plot the output - note that the plot range and the x axis have to be adjusted based on the length of the table
# determine x axis range 
x_range <- 0:(length(output[,1])-1)
# this plots the total population size
plot(x_range,output[,1]+output[,2],type='l',ylim=c(0,max(output[,1]+output[,2])),xlab = "Generation",ylab = "Population size")
# add number of wild type individuals
lines(x_range,output[,1], col="blue")
# add number of mutant individuals
lines(x_range,output[,2], col="red")
```




## Simulating a whole population

Generate a population with the influence of events. 
Here we are trying to determine the right parameters to then fix and later only vary with the frequency (t_pause).

```{r}
t_pause = 1 #t_switch, time between events
t_event = 5 #duration of event (Working: 5)
t_switch = t_pause
sel_coeff = 0.0

#flooding_table <- c()

simulate_pop <- function(N_init_a, N_init_A, decay_rate, sel_coeff, mut_rate, t_max, t_pause) {
  # Create the vector in which to save the results
  pop_vector <- c(N_init_a, N_init_A)
  # initiate the variables
  pop_new <- c(N_init_a, N_init_A)
  
  ph = 0 #phase in the beginning of the simulation

  
  # run the simulation until generation t_max
  for (i in 1:t_max+1) {
    if(ph %%2 == 0){ #checking in which phase the simulation is, still needs fine-tuning
    decay_rate = -0.02 #no event - parameters
    sel_coeff = -0.015 #(Working: -0.02 & -0.015)
  }else{
    decay_rate = 0.03 #event - parameters
    sel_coeff = 0.05 #(Working: 0.03 & 0.05)
  }

  if (i >= t_switch) { #changing the phase number, if the simulation is at the end of the phase
    if(ph %% 2 == 0) {
      t_switch = t_switch + t_event
      ph = ph + 1
    }else{
      t_switch = t_switch + t_pause
      ph = ph + 1
    }
  }
    
    # redefine the current population one generation later
    pop_new <- simulate_one_gen(pop_new[1],pop_new[2], decay_rate, sel_coeff, mut_rate)
    # add the new population sizes to the output vector
    pop_vector <- rbind(pop_vector,pop_new)
    # condition to stop the simulation before t_max: either the population exceeds 1.5 times the original population size, or it goes extinct
    if (pop_new[1]+pop_new[2]>=1.5*(N_init_a+ N_init_A) | pop_new[1]+pop_new[2]==0) break
  }
  
  # define the row and column names of the output vector
  rownames(pop_vector) <- (0:t_max)[1:length(pop_vector[,1])] # note that the vector has to be cut if the simulation stopped early
  colnames(pop_vector) <- c("WT","Mut")
  # return the result
  return(pop_vector)	
}

# Test the function and plot the result
# set t_max to a value to be used throughout
max_gen <- 1000
# create your simulation data
output <- simulate_pop(1000,0,0.35,0.01,0.001,max_gen, t_pause) #(Working: 1000, 0, 0.35, 0.01, 0.001, max_gen)
# show the last few lines of the data table
print(output)
# plot the output - note that the plot range and the x axis have to be adjusted based on the length of the table
# determine x axis range 
x_range <- 0:(length(output[,1])-1)
# this plots the total population size
plot(x_range,output[,1]+output[,2],type='l',ylim=c(0,max(output[,1]+output[,2])),xlab = "Generation",ylab = "Population size")
# add number of wild type individuals
lines(x_range,output[,1], col="blue")
# add number of mutant individuals
lines(x_range,output[,2], col="red")

```


##  Generating multiple replicates and Analyze data
To calculate the probability of rescue & find the means of the population sizes

```{r}
# Set some parameters to fixed values
init_a <- 1000
init_A <- 0
m_rate <- 0.001
max_gen <- 1000
no_replicates <- 100

# Frequencies to vary
frequencies <- c(1:10)

# Initialize data tables to collect results
data_table_Pop <- c()
data_table_Nw <- c()
data_table_Nm <- c()
data_table_comp <- c()

# Initialize counters for survival
surviving_Nw <- numeric(length(frequencies))
surviving_Nm <- numeric(length(frequencies))

# Run the simulation across all chosen parameters
for (t_pause in frequencies) {
  i <- 1
  repeat {
    # Run the simulation once
    one_run <- simulate_pop(init_a, init_A, 0, 0, m_rate, max_gen, t_pause)
    
    # Determine the final sizes of the populations
    size_wt <- one_run[nrow(one_run), 1]
    size_mut <- one_run[nrow(one_run), 2]
    
    # Count surviving populations
    if (size_wt > 0) {
      surviving_Nw[t_pause] <- surviving_Nw[t_pause] + 1
    }
    if (size_mut > 0) {
      surviving_Nm[t_pause] <- surviving_Nm[t_pause] + 1
    }

    # Enter the data into the tables
    final_size <- size_wt + size_mut
    final_gen <- nrow(one_run)
    
    data_table_Pop <- rbind(data_table_Pop, c(t_pause, final_size, final_gen))
    data_table_Nw <- rbind(data_table_Nw, c(t_pause, size_wt, final_gen))
    data_table_Nm <- rbind(data_table_Nm, c(t_pause, size_mut, final_gen))
    
    portion_WT <- size_wt / final_size
    portion_Mut <- size_mut / final_size
    
    data_table_comp <- rbind(data_table_comp, c(t_pause, portion_WT, portion_Mut))
    
    # Stop after the specified number of replicates
    if (i >= no_replicates) break
    i <- i + 1
  }
}

# Define column names of data tables
colnames(data_table_Pop) <- c("f", "Final Pop", "final_gen")
colnames(data_table_Nw) <- c("f", "Final WT", "final_gen")
colnames(data_table_Nm) <- c("f", "Final Mut", "final_gen")
colnames(data_table_comp) <- c("f", "Proportion_WT", "Proportion_Mut")

# Calculate ratios of surviving populations
ratios <- data.frame(
  t_pause = frequencies,
  Nw_survival_ratio = surviving_Nw / no_replicates,
  Nm_survival_ratio = surviving_Nm / no_replicates
)

#plot(ratios[,1], ratios[,2], xlab = "Frequencies of events", ylab = "Survival-ratio", main = "Survival-ratio of WT", col = "chocolate", type = "p", )
#lines(ratios[,1], ratios[,3], col = "darkolivegreen4", type = "l")

```

## Plottig the data
```{r}
# Ensure ratios are numeric
ratios$t_pause <- as.numeric(as.character(ratios$t_pause))
ratios$Nw_survival_ratio <- as.numeric(as.character(ratios$Nw_survival_ratio))
ratios$Nm_survival_ratio <- as.numeric(as.character(ratios$Nm_survival_ratio))

# Plot for Wildtype survival ratio
#par(mfrow = c(1,2))

plot(ratios$t_pause, ratios$Nw_survival_ratio, 
     xlab = "Pause between Events", 
     ylab = "Survival Ratio", 
     main = c(paste("Survival Ratio of wildtype and mutant"), paste("depending on the frequency of events")), 
     col = "chocolate", 
     type = "b",
     pch = c(16))  # Use a solid circle for points

# Add lines for Mutation survival ratio
#plot(ratios$t_pause, ratios$Nm_survival_ratio, 
      #col = "darkolivegreen4", 
      #type = "b")  # Use points and lines

# Optionally, add a legend
legend("bottomright", legend = c("Wildtype"), 
       col = c("chocolate"),
       pch = c(16, NA))

```

```{r}
mean_comp_tot <- c()

mean_comp_WT <- aggregate(data_table_comp[,2]~ f, data = data_table_comp, FUN = mean)
mean_comp_Mut <- aggregate(data_table_comp[,3] ~ f, data = data_table_comp, FUN = mean)
mean_comp_tot <- rbind(mean_comp_tot, c(mean_comp_WT, mean_comp_Mut))

# Load necessary libraries
library(ggplot2)
library(dplyr)
library(tidyr)

# Combine the two data frames into one
# Ensure the names are consistent for merging
colnames(mean_comp_WT) <- c("f", "Proportion_WT")
colnames(mean_comp_Mut) <- c("f", "Proportion_Mut")

# Merge the data
mean_comp <- merge(mean_comp_WT, mean_comp_Mut, by = "f")

# Reshape the data into long format for ggplot2
mean_comp_long <- mean_comp %>%
  pivot_longer(cols = c("Proportion_WT", "Proportion_Mut"), 
               names_to = "Type", 
               values_to = "Proportion")

# Plot the stacked bar chart
ggplot(mean_comp_long, aes(x = as.factor(f), y = Proportion, fill = Type)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(x = "Pause between Events", 
       y = "Mean Proportion", 
       title = "Mean Proportion of Wildtype and Mutant by Event Frequency") +
  scale_fill_manual(values = c("Proportion_WT" = "chocolate", "Proportion_Mut" = "darkolivegreen4"),
                    labels = c("Mutant","Wildtype")) +
  theme_minimal()


#library(ggplot2)

#ggplot(data_table_comp, aes(fill=condition, y=data_table_comp$, x=specie)) + 
    #geom_bar(position="dodge", stat="identity")

```
```{r}
#data_table_Pop
# Ensure data_table_Pop is a data frame
data_table_Pop <- as.data.frame(data_table_Pop)

# Set column names (if not already set)
colnames(data_table_Pop) <- c("f", "Final_Pop", "final_gen")

# Initialize a vector to store extinction ratios for each frequency
extinction_ratios <- numeric(length(frequencies))

# Calculate extinction ratios for each frequency
for (t_pause in frequencies) {
  # Filter data for the current frequency
  current_data <- subset(data_table_Pop, f == t_pause)
  
  # Count how many populations went extinct (Final Pop == 0)
  extinct_count <- sum(current_data$Final_Pop == 0)
  
  # Calculate the extinction ratio (extinct_count / no_replicates)
  extinction_ratios[t_pause] <- extinct_count / no_replicates
}

# Create a data frame for extinction ratios
extinction_df <- data.frame(
  t_pause = frequencies,
  extinction_ratio = extinction_ratios
)

# Print or view the extinction ratios
print(extinction_df)

plot(extinction_df[,1], extinction_df[,2], 
     type = "b", 
     pch = 16, 
     col = "darkorchid3",
     ylab = "Extinction Probability", 
     xlab = "Pause between Events",
     main = "Extinction rate depending on frequency")

```

```{r}
ggplot(data_table_Pop, aes(x = as.factor(f), y = final_gen)) +
  geom_boxplot(fill = "darkgoldenrod1") +
  labs(x = "Pause between Events", y = "Final Generation Count", 
       title = "Distribution of Final Generations by Pause between Events") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))  # Center the title


```
```{r}
ggplot(data_table_Pop, aes(x = final_gen, fill = as.factor(f))) +
  geom_histogram(binwidth = 10, position = "identity", alpha = 0.6) +
  labs(x = "Final Generation Count", y = "Count", 
       title = "Histogram of Final Generations by Pause between Events") +
  theme_minimal() +
  scale_fill_discrete(name = "Pause btw. Events") +
  theme(plot.title = element_text(hjust = 0.5))  # Center the title

```

```{r}

# Filter for populations that went extinct (Final_Pop == 0)
extinct_pop <- subset(data_table_Pop, Final_Pop == 0)

# Select only the frequency and generation of extinction
extinct_data <- extinct_pop[, c("f", "final_gen")]

# Optionally, rename columns for clarity
colnames(extinct_data) <- c("Frequency", "Extinction_Generation")

# View or print the resulting data
print(extinct_data)

#Plot the data
ggplot(extinct_data, aes(x=Frequency, y=Extinction_Generation)) +
       geom_point(col = "aquamarine4", pch = 19) +
  xlim(c(1,10)) +
  geom_smooth(method=lm , color="coral", se=FALSE) + 
  labs(title = "Generation of Extinction across different Frequencies", x = "Pause between Events", y = "Extinction Generation") +
  theme_bw() +
  theme(
    legend.position = c(.95, .95),
    legend.justification = c("right", "top"),
    legend.box.just = "right",
    legend.margin = margin(6, 6, 6, 6)
    ) +
  theme(plot.title = element_text(hjust = 0.5)) # Center the title
  





```
```{r}
# Numbers to allocate
numbers <- 1:6

# Randomly shuffle the numbers
shuffled_numbers <- sample(numbers)

# Split the numbers equally between person A and person B
person_a <- shuffled_numbers[1:3]
person_b <- shuffled_numbers[4:6]

# Print the results
cat("Person A gets:", person_a, "\n")
cat("Person B gets:", person_b, "\n")

```


## Test area:
Code from the "Test Area" (just some coding bits that were used to try to figure stuff out, nothing really worked) and the part about reproducability were moved, for the sake of tidiness and better overview, to a different document (THEE-Project-Appendix).
