---
title: "Evolutionary rescue in flooding events"
author: "Ramin Dürst & Matthias Schönholzer"
date: "16.10.2024"
output:
  html_document:
    df_print: paged
co-author: "Base code by Claudia Bank"
---

## List of variables and parameters

-   $N_a$ current size of the wild-type (a) population (changes over time)
-   $N_A$ current size of the mutant (A) population (changes over time)
-   `N_init_a` initial population size of wild-type population (should be a positive integer)
-   `N_init_A` initial population size of the mutant population (~~should be a non-negative integer~~ –\> will be 0 at start)
-   `decay_rate` detrimental effect of the environment on the wild type (i.e., the wild-type population decays at this rate; should be \>0 to induce potential extinction)
-   `sel_coeff` ~~selective advantage of the mutant over the wild type (should be \>decay_rate to allow for evolutionary rescue)~~ –\> selection coefficient of wild type declines during the event, while the selection coefficient of the mutant increases during an event.
-   `mut_rate` mutation rate from the wild type to the mutant genotype (should be \>0 and small)
-   `t_max` number of generations after which to stop the simulation (should be an integer \>0; this is not a model parameter but a setting for the simulations). In the code presented here, the simulation stops automatically if the population goes extinct, or if it recovers and exceeds 1.5 times the original total population size, so `t_max` can be set to a large value without creating extremely long simulation times.
-   `duration of events` the timespan of generations the event occurs. Depends on the decay rate of the wild type and will be over 1.
-   `frequency of events` at which frequency the events occur over time
-   `strength of events` how strong the events affect the population. Is a combination of decay / growth rate and selection coefficient.

## Scenario / Biological question

A hypothetical population of plants in a water meadow by a river with two genotypes: one which does bad during flooding (wildtype) and one which does good during flooding (mutant). The frequency of being flooded increases due to climate change and depending on frequency, either the wildtype persists, the mutant takes over, or both genotypes and therefor the population go extinct.

How does the increasing frequency of flooding events affect the wildtype and mutant population of a water meadow?


## Simulate one generation of population size change

We want to observe the numbers of wild-type and mutant individuals over time. As first step, we simulate one generation of the population. At time $t$, the population has $N_a$ wild-type individuals and $N_A$ mutant individuals. First, these reproduce. This is modeled by drawing a Poisson random variable according to the expected absolute number $N_a \cdot (1-r)$ of wild-type individuals, where $r$ is the decay rate of the wild type, and $N_A \cdot (1-r+s)$ of mutant individuals, where $s$ is the selective advantage of the mutant over the wild time. Then mutation from $a$ to $A$ occurs as a Poisson random variable according to the given mutation rate. This defines the composition of the population at time $t+1$.

```{r}
#Original

simulate_one_gen <- function(N_a, N_A, decay_rate, sel_coeff, mut_rate) {
  # draw offspring according to Poisson distribution
  offsp_a <- rpois(1, N_a * (1-decay_rate))
  offsp_A <- rpois(1, N_A * (1-decay_rate+sel_coeff))
  # draw new mutants according to Poisson distribution
  mut_a_to_A <- rpois(1, offsp_a * mut_rate)
  
  # determine new population sizes of wild type and mutant
  N_a_new <- max(offsp_a - mut_a_to_A, 0)
  N_A_new <-  offsp_A + mut_a_to_A
  
  return(c(N_a_new, N_A_new))
}
# Test the function
print(simulate_one_gen(100,0,0.1,0.2,0.001))
```

## Simulate a population trajectory

We can now track the dynamics of the population from one generation to another, but this is not enough. We want to follow the population from time 0 until it either is rescued or goes extinct. This is what the following function does.

```{r}
#Original

simulate_pop <- function(N_init_a, N_init_A, decay_rate, sel_coeff, mut_rate, t_max) {
  # Create the vector in which to save the results
  pop_vector <- c(N_init_a, N_init_A)
  # initiate the variables
  pop_new <- c(N_init_a, N_init_A)
  
  # run the simulation until generation t_max
  for (i in 1:t_max+1) {
    # redefine the current population one generation later
    pop_new <- simulate_one_gen(pop_new[1],pop_new[2], decay_rate, sel_coeff, mut_rate)
    # add the new population sizes to the output vector
    pop_vector <- rbind(pop_vector,pop_new)
    # condition to stop the simulation before t_max: either the population exceeds 1.5 times the original population size, or it goes extinct
    if (pop_new[1]+pop_new[2]>=1.5*(N_init_a+ N_init_A) | pop_new[1]+pop_new[2]==0) break
  }
  
  # define the row and column names of the output vector
  rownames(pop_vector) <- (0:t_max)[1:length(pop_vector[,1])] # note that the vector has to be cut if the simulation stopped early
  colnames(pop_vector) <- c("a","A")
  # return the result
  return(pop_vector)	
}


# Test the function and plot the result
# set t_max to a value to be used througout
max_gen <- 1000
# create your simulation data, without mut_rate & sel_coeff and a higher decay_rate
output <- simulate_pop(1000,0,0.1,0.2,0.001,max_gen)
# show the last few lines of the data table
print(output)
# plot the output - note that the plot range and the x axis have to be adjusted based on the length of the table
# determine x axis range 
x_range <- 0:(length(output[,1])-1)
# this plots the total population size
plot(x_range,output[,1]+output[,2],type='l',ylim=c(0,max(output[,1]+output[,2])),xlab = "Generation",ylab = "Population size")
# add number of wild type individuals
lines(x_range,output[,1], col="blue")
# add number of mutant individuals
lines(x_range,output[,2], col="red")
```




## Simulating a whole population

Generate a population with the influence of events. 
Here we are trying to determine the right parameters to then fix and later only vary with the frequency (t_pause).

```{r}
t_pause = 1 #t_switch, time between events
t_event = 5 #duration of event (Working: 5)
t_switch = t_pause
sel_coeff = 0.0

#flooding_table <- c()

simulate_pop <- function(N_init_a, N_init_A, decay_rate, sel_coeff, mut_rate, t_max, t_pause) {
  # Create the vector in which to save the results
  pop_vector <- c(N_init_a, N_init_A)
  # initiate the variables
  pop_new <- c(N_init_a, N_init_A)
  
  ph = 0 #phase in the beginning of the simulation

  
  # run the simulation until generation t_max
  for (i in 1:t_max+1) {
    if(ph %%2 == 0){ #checking in which phase the simulation is, still needs fine-tuning
    decay_rate = -0.02 #no event - parameters
    sel_coeff = -0.015 #(Working: -0.02 & -0.015)
  }else{
    decay_rate = 0.03 #event - parameters
    sel_coeff = 0.05 #(Working: 0.03 & 0.05)
  }

  if (i >= t_switch) { #changing the phase number, if the simulation is at the end of the phase
    if(ph %% 2 == 0) {
      t_switch = t_switch + t_event
      ph = ph + 1
    }else{
      t_switch = t_switch + t_pause
      ph = ph + 1
    }
  }
    
    # redefine the current population one generation later
    pop_new <- simulate_one_gen(pop_new[1],pop_new[2], decay_rate, sel_coeff, mut_rate)
    # add the new population sizes to the output vector
    pop_vector <- rbind(pop_vector,pop_new)
    # condition to stop the simulation before t_max: either the population exceeds 1.5 times the original population size, or it goes extinct
    if (pop_new[1]+pop_new[2]>=1.5*(N_init_a+ N_init_A) | pop_new[1]+pop_new[2]==0) break
  }
  
  # define the row and column names of the output vector
  rownames(pop_vector) <- (0:t_max)[1:length(pop_vector[,1])] # note that the vector has to be cut if the simulation stopped early
  colnames(pop_vector) <- c("WT","Mut")
  # return the result
  return(pop_vector)	
}

# Test the function and plot the result
# set t_max to a value to be used throughout
max_gen <- 1000
# create your simulation data
output <- simulate_pop(1000,0,0.35,0.01,0.001,max_gen, t_pause) #(Working: 1000, 0, 0.35, 0.01, 0.001, max_gen)
# show the last few lines of the data table
print(tail(output))
# plot the output - note that the plot range and the x axis have to be adjusted based on the length of the table
# determine x axis range 
x_range <- 0:(length(output[,1])-1)
# this plots the total population size
plot(x_range,output[,1]+output[,2],type='l',ylim=c(0,max(output[,1]+output[,2])),xlab = "Generation",ylab = "Population size")
# add number of wild type individuals
lines(x_range,output[,1], col="blue")
# add number of mutant individuals
lines(x_range,output[,2], col="red")

```


## Generating multiple replicates at different frequencies and extracting the data

After generating a single population with a given frequency (t_event), we now want to create replicates at multiple frequencies. We then create different data-tables which contain different values and can be used for further calculations and plots:

-   data_table_Pop: frequency, final total population, final generation

-   data_table_Nw: frequency, final population of wildtype, final generation

-   data_table_Nm: frequency, final population of mutant, final generation

-   data_table_comp: frequency, proportion of wildtype, proportion of mutant

While the first three data-tables give the final population of either the total, wildtype, or mutant and the corresponding generation depending on the frequency, the fourth data-table contains the proportions of wildtype and mutant depending on the frequency.

Additionally, we count the number of wildtype- and mutant-populations which survived after the set conditions and calculated the ratio of surviving.

```{r}
# Set some parameters to fixed values
init_a <- 1000
init_A <- 0
m_rate <- 0.001
max_gen <- 1000
no_replicates <- 100

# Frequencies to vary
frequencies <- c(1:10)

# Initialize data tables to collect results
data_table_Pop <- c()
data_table_Nw <- c()
data_table_Nm <- c()
data_table_comp <- c()

# Initialize counters for survival
surviving_Nw <- numeric(length(frequencies))
surviving_Nm <- numeric(length(frequencies))

# Run the simulation across all chosen parameters
for (t_pause in frequencies) {
  i <- 1
  repeat {
    # Run the simulation once
    one_run <- simulate_pop(init_a, init_A, 0, 0, m_rate, max_gen, t_pause)
    
    # Determine the final sizes of the populations
    size_wt <- one_run[nrow(one_run), 1]
    size_mut <- one_run[nrow(one_run), 2]
    
    # Count surviving populations
    if (size_wt > 0) {
      surviving_Nw[t_pause] <- surviving_Nw[t_pause] + 1
    }
    if (size_mut > 0) {
      surviving_Nm[t_pause] <- surviving_Nm[t_pause] + 1
    }

    # Enter the data into the tables
    final_size <- size_wt + size_mut
    final_gen <- nrow(one_run)
    
    data_table_Pop <- rbind(data_table_Pop, c(t_pause, final_size, final_gen))
    data_table_Nw <- rbind(data_table_Nw, c(t_pause, size_wt, final_gen))
    data_table_Nm <- rbind(data_table_Nm, c(t_pause, size_mut, final_gen))
    
    portion_WT <- size_wt / final_size
    portion_Mut <- size_mut / final_size
    
    data_table_comp <- rbind(data_table_comp, c(t_pause, portion_WT, portion_Mut))
    
    # Stop after the specified number of replicates
    if (i >= no_replicates) break
    i <- i + 1
  }
}

# Define column names of data tables
colnames(data_table_Pop) <- c("f", "Final Pop", "final_gen")
colnames(data_table_Nw) <- c("f", "Final WT", "final_gen")
colnames(data_table_Nm) <- c("f", "Final Mut", "final_gen")
colnames(data_table_comp) <- c("f", "Proportion_WT", "Proportion_Mut")

# Calculate ratios of surviving populations
ratios <- data.frame(
  t_pause = frequencies,
  Nw_survival_ratio = surviving_Nw / no_replicates,
  Nm_survival_ratio = surviving_Nm / no_replicates
)

#plot(ratios[,1], ratios[,2], xlab = "Frequencies of events", ylab = "Survival-ratio", main = "Survival-ratio of WT", col = "chocolate", type = "p", )
#lines(ratios[,1], ratios[,3], col = "darkolivegreen4", type = "l")

```

## Plottig the data

# Plotting the survival-ratio of the wildtype

Here we plot the survival-ratio of the wildtype depending on the pause between events according to the data we got above. The higher the pause between events, the lower the frequency of events. The graph shows that the less frequent events happen (the higher the pause between events is), the higher is the probability of survival.

```{r}
# Ensure ratios are numeric
ratios$t_pause <- as.numeric(as.character(ratios$t_pause))
ratios$Nw_survival_ratio <- as.numeric(as.character(ratios$Nw_survival_ratio))
ratios$Nm_survival_ratio <- as.numeric(as.character(ratios$Nm_survival_ratio))

# Plot for Wildtype survival ratio
#par(mfrow = c(1,2))

plot(ratios$t_pause, ratios$Nw_survival_ratio, 
     xlab = "Pause between Events", 
     ylab = "Survival Ratio", 
     main = c(paste("Survival Ratio of wildtype"), paste("depending on the frequency of events")), 
     col = "chocolate", 
     type = "b",
     pch = c(16))  # Use a solid circle for points

# Add lines for Mutation survival ratio
#plot(ratios$t_pause, ratios$Nm_survival_ratio, 
      #col = "darkolivegreen4", 
      #type = "b")  # Use points and lines

# Optionally, add a legend
legend("bottomright", legend = c("Wildtype"), 
       col = c("chocolate"),
       pch = c(16, NA))

```

# Mean proportion of wildtype and mutant according to the frequency

We want to show the composition of the population at different frequencies. In the first step, we create two data-tables with the frequencies and the corresponding average proportion of the mutant and the wildtype. The values are extracted from the data_table_comp created under "Generating multiple replicates at different frequencies and extracting the data". The two data-tables are bind together into one table, with which we create the stacked barplot. 
The plot shows that populations at high frequencies (low pauses) contain mainly mutants, while a low frequency (high pause) results in populations mainly compromised of wildtypes.

```{r}
mean_comp_tot <- c()

mean_comp_WT <- aggregate(data_table_comp[,2]~ f, data = data_table_comp, FUN = mean)
mean_comp_Mut <- aggregate(data_table_comp[,3] ~ f, data = data_table_comp, FUN = mean)
mean_comp_tot <- rbind(mean_comp_tot, c(mean_comp_WT, mean_comp_Mut))

# Load necessary libraries
library(ggplot2)
library(dplyr)
library(tidyr)

# Combine the two data frames into one
# Ensure the names are consistent for merging
colnames(mean_comp_WT) <- c("f", "Proportion_WT")
colnames(mean_comp_Mut) <- c("f", "Proportion_Mut")

# Merge the data
mean_comp <- merge(mean_comp_WT, mean_comp_Mut, by = "f")

# Reshape the data into long format for ggplot2
mean_comp_long <- mean_comp %>%
  pivot_longer(cols = c("Proportion_WT", "Proportion_Mut"), 
               names_to = "Type", 
               values_to = "Proportion")

# Plot the stacked bar chart
ggplot(mean_comp_long, aes(x = as.factor(f), y = Proportion, fill = Type)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(x = "Pause between Events", 
       y = "Mean Proportion", 
       title = "Mean Proportion of Wildtype and Mutant by Event Frequency") +
  scale_fill_manual(values = c("Proportion_WT" = "chocolate", "Proportion_Mut" = "darkolivegreen4"),
                    labels = c("Mutant","Wildtype")) +
  theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5))  # Center the title



#library(ggplot2)

#ggplot(data_table_comp, aes(fill=condition, y=data_table_comp$, x=specie)) + 
    #geom_bar(position="dodge", stat="identity")

```




#trying out some stuff (namely plotting just the WT)
```{r}
mean_comp_WT_long <- mean_comp %>%
  pivot_longer(cols = "Proportion_WT", 
               names_to = "Type", 
               values_to = "Proportion")
ggplot(mean_comp_WT_long, aes(x = as.factor(f), y = Proportion, fill = Type)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(x = "Pause between Events", 
       y = "Mean Proportion", 
       title = "Mean Proportion of Wildtype by Event Frequency") +
  scale_fill_manual(values = ("Proportion_WT" = "chocolate"),
                    labels = ("Wildtype")) +
  theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5))
```



# Plot for Exctinction rate depending of the Frequency of events

Here we want to see, how the probability of the population going extinct changes with different frequencies of events. We extract the data for extinction from the large data table (data_table_Pop) by finding the ratio of times the population went extinct to the total amount of replicates. This ratio is then visualized in a plot.

The plot shows that with increasing time between each event decreases the probability that the population goes extinct. 

```{r}
#data_table_Pop
# Ensure data_table_Pop is a data frame
data_table_Pop <- as.data.frame(data_table_Pop)

# Set column names (if not already set)
colnames(data_table_Pop) <- c("f", "Final_Pop", "final_gen")

# Initialize a vector to store extinction ratios for each frequency
extinction_ratios <- numeric(length(frequencies))

# Calculate extinction ratios for each frequency
for (t_pause in frequencies) {
  # Filter data for the current frequency
  current_data <- subset(data_table_Pop, f == t_pause)
  
  # Count how many populations went extinct (Final Pop == 0)
  extinct_count <- sum(current_data$Final_Pop == 0)
  
  # Calculate the extinction ratio (extinct_count / no_replicates)
  extinction_ratios[t_pause] <- extinct_count / no_replicates
}

# Create a data frame for extinction ratios
extinction_df <- data.frame(
  t_pause = frequencies,
  extinction_ratio = extinction_ratios
)

# Print or view the extinction ratios
print(extinction_df)

plot(extinction_df[,1], extinction_df[,2], 
     type = "b", 
     pch = 16, 
     col = "darkorchid3",
     ylab = "Extinction Probability", 
     xlab = "Pause between Events",
     main = "Extinction rate depending on frequency")

```
#trying stuff out (namely plotting extinction rate of MUT and WT separately)
```{r}
par(mfrow = c(1,3))
#data_table_Pop
# Ensure data_table_Pop is a data frame
data_table_Pop <- as.data.frame(data_table_Pop)

# Set column names (if not already set)
colnames(data_table_Pop) <- c("f", "Final_Pop", "final_gen")

# Initialize a vector to store extinction ratios for each frequency
extinction_ratios <- numeric(length(frequencies))

# Calculate extinction ratios for each frequency
for (t_pause in frequencies) {
  # Filter data for the current frequency
  current_data <- subset(data_table_Pop, f == t_pause)
  
  # Count how many populations went extinct (Final Pop == 0)
  extinct_count <- sum(current_data$Final_Pop == 0)
  
  # Calculate the extinction ratio (extinct_count / no_replicates)
  extinction_ratios[t_pause] <- extinct_count / no_replicates
}

# Create a data frame for extinction ratios
extinction_df <- data.frame(
  t_pause = frequencies,
  extinction_ratio = extinction_ratios
)

# Print or view the extinction ratios
print(extinction_df)

plot(extinction_df[,1], extinction_df[,2], 
     type = "b", 
     pch = 16, 
     col = "darkorchid3",
     ylab = "Extinction Probability Tot-Pop", 
     xlab = "Pause between Events",
     main = "Tot-Pop-Extinction rate dep. frequency")

#data_table_Pop
# Ensure data_table_Pop is a data frame
data_table_Nw <- as.data.frame(data_table_Nw)

# Set column names (if not already set)
colnames(data_table_Nw) <- c("f", "Final_Pop", "final_gen")

# Initialize a vector to store extinction ratios for each frequency
extinction_ratios_WT <- numeric(length(frequencies))

# Calculate extinction ratios for each frequency
for (t_pause in frequencies) {
  # Filter data for the current frequency
  current_data_WT <- subset(data_table_Nw, f == t_pause)
  
  # Count how many populations went extinct (Final Pop == 0)
  extinct_count_WT <- sum(current_data_WT$Final_Pop == 0)
  
  # Calculate the extinction ratio (extinct_count / no_replicates)
  extinction_ratios_WT[t_pause] <- extinct_count_WT / no_replicates
}

# Create a data frame for extinction ratios
extinction_WT_df <- data.frame(
  t_pause = frequencies,
  extinction_ratio_WT = extinction_ratios_WT
)

# Print or view the extinction ratios
print(extinction_WT_df)

plot(extinction_WT_df[,1], extinction_WT_df[,2], 
     type = "b", 
     pch = 16, 
     col = "chartreuse4",
     ylab = "Extinction Probability WT", 
     xlab = "Pause between Events",
     main = "WT-Extinction rate dep. frequency")


#data_table_Pop
# Ensure data_table_Pop is a data frame
data_table_Nm <- as.data.frame(data_table_Nm)

# Set column names (if not already set)
colnames(data_table_Nm) <- c("f", "Final_Pop", "final_gen")

# Initialize a vector to store extinction ratios for each frequency
extinction_ratios_MUT <- numeric(length(frequencies))

# Calculate extinction ratios for each frequency
for (t_pause in frequencies) {
  # Filter data for the current frequency
  current_data_MUT <- subset(data_table_Nm, f == t_pause)
  
  # Count how many populations went extinct (Final Pop == 0)
  extinct_count_MUT <- sum(current_data_MUT$Final_Pop == 0)
  
  # Calculate the extinction ratio (extinct_count / no_replicates)
  extinction_ratios_MUT[t_pause] <- extinct_count_MUT / no_replicates
}

# Create a data frame for extinction ratios
extinction_MUT_df <- data.frame(
  t_pause = frequencies,
  extinction_ratio_MUT = extinction_ratios_MUT
)

# Print or view the extinction ratios
print(extinction_MUT_df)

plot(extinction_MUT_df[,1], extinction_MUT_df[,2], 
     type = "b", 
     pch = 16, 
     col = "brown3",
     ylab = "Extinction Probability MUT", 
     xlab = "Pause between Events",
     main = "MUT-Extinction rate dep. frequency")
par(mfrow= c(1,1))
```



# Plot of Distribution of final Generations depending on the Frequency

Here we want to see the relationship between the pause between the events and the number of generations, the population was simulated. The conditions for the simulation to end were either extinction or the population growing to a size of 1.5 times the starting population.
In this plot it can be seen, that the populations at medium frequencies ran for slightly longer. 
This can be explained with the following theories: Populations with short pauses either die out quickly or grow fast. There is a aspect of stochasticity with the mutation rate. At longer pauses between the events the mutation cannot fix and the wildtype then grows to a large population. At the intermediate pauses the population struggles more to  grow to the stopping condition. There are to few events for the mutation to take over and to few for the wildtype to just grow without problems. 

```{r}
ggplot(data_table_Pop, aes(x = as.factor(f), y = final_gen)) +
  geom_boxplot(fill = "darkgoldenrod1") +
  labs(x = "Pause between Events", y = "Final Generation Count", 
       title = "Distribution of Final Generations by Pause between Events") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))  # Center the title


```

# Plot of Counts of final Generations depending on the Frequency (Histogram)

Similar to the plot before, we want to visualize the relationship of the final generations and the frequencies of the events. The plot is not very clear, but it shows the same effect. On the higher end of generations are the populations with intermediate pauses (green-blueish color). Longer pauses show shorter simulation runs (pink color).

```{r}
ggplot(data_table_Pop, aes(x = final_gen, fill = as.factor(f))) +
  geom_histogram(binwidth = 10, position = "identity", alpha = 0.6) +
  labs(x = "Final Generation Count", y = "Count", 
       title = "Histogram of Final Generations by Pause between Events") +
  theme_minimal() +
  scale_fill_discrete(name = "Pause btw. Events") +
  theme(plot.title = element_text(hjust = 0.5))  # Center the title

```

# Plotting the extinction-generation across different frequencies

At last we wanted to show the differences of the generation of extinction depending on the frequency. To do this, we extracted the replicates which went extinct from the data-table from the whole population. Then we created a table which contains the corresponding frequencies and the number of the final generation and created a scatterplot. We included the a trend line, which indicates the mean of the frequencies. As can be seen in the plot, at certain frequencies there are no datapoints, which means at these frequencies the population did not go extinct throughout all replicates.

```{r}

# Filter for populations that went extinct (Final_Pop == 0)
extinct_pop <- subset(data_table_Pop, Final_Pop == 0)

# Select only the frequency and generation of extinction
extinct_data <- extinct_pop[, c("f", "final_gen")]

# Optionally, rename columns for clarity
colnames(extinct_data) <- c("Frequency", "Extinction_Generation")

# View or print the resulting data
print(extinct_data)

#Plot the data
ggplot(extinct_data, aes(x=Frequency, y=Extinction_Generation)) +
       geom_point(col = "aquamarine4", pch = 19) +
  geom_smooth(method=lm , color="coral", se=FALSE) + 
  labs(title = "Generation of Extinction across different Frequencies", x = "Pause between Events", y = "Extinction Generation") +
  theme_bw() +
  theme(
    legend.position = c(.95, .95),
    legend.justification = c("right", "top"),
    legend.box.just = "right",
    legend.margin = margin(6, 6, 6, 6)
    ) +
  theme(plot.title = element_text(hjust = 0.5)) + # Center the title
  scale_x_continuous(breaks = 1:10, limits = c(1,10)) # Set x-axis to show only natural numbers 1 through 10





```

## Biological explanations

Our model simulates a simple scenario, where a plant population in a floodplain near a river experiences regular flooding events. The frequencies of these flooding increases due to climate change. The flooding of the plains is detrimental for the wildtype population, but there is a mutation which would be beneficial in case of flooding and benefits during events. 
We want to simulate this scenario with different frequencies of events and see how the population changes. 
For this, we calculated and plotted multiple parameters and variables to show these relationships. 

From our model we can deduce, that the frequency of flooding events has a large impact on the fate and composition of the population. With increasing frequency/shorter pauses between events, the mutation becomes more beneficial and rescues the population. At lower frequencies/long pauses the wildtype stays dominant in the population and persists.

