---
title: "THEE Project - Appendix"
output: html_notebook
---
## Making your results reproducible

When you generate “random numbers” in R, you are actually generating pseudorandom numbers. These numbers are generated with an algorithm that requires a seed to initialize. This means that if you know the seed and the generator, you can reproduce the output.

```{r}
# Draw 5 (pseudo)random numbers from a normal distribution
rnorm(5)
# Draw again 5 (pseudo)random numbers from a normal distribution
rnorm(5)
```

Expectedly, the numbers are different. How can we reproduce the exact output?

```{r}
#initialize the seed of the pseudo random generator
set.seed(1)
# Draw 5 (pseudo)random numbers from a normal distribution
rnorm(5)
#initialize the seed of the pseudo random generator
set.seed(1)
# Draw again 5 (pseudo)random numbers from a normal distribution
rnorm(5)
```


## Test-Area
This are just bits of code, that were used for the THEE project, to test things out. Nothing really worked as intended. It is here for the sake of hoarding and out of fear of deleting important bits.

```{r}
# Parameters
decay_rate = c(0.35, 1.1) # decay rate in presence and absence of event
freq_event = 10 # every 10 generations an event happens
dur_event = 2 # event has a duration of 2 generations
t_max = 100
gen = 0

# Create a list to store the results of each simulation
simulation_results <- list()

# Function to simulate one generation
simulate_one_gen_2 <- function(N_a, N_A, decay_rate, growth_rate, sel_coeff, mut_rate) {
  offsp_a <- rpois(1, N_a * (1 - decay_rate))
  offsp_A <- rpois(1, N_A * (1 - decay_rate + sel_coeff))
  
  # Draw new mutants according to Poisson distribution
  mut_a_to_A <- rpois(1, offsp_a * mut_rate)
  
  # Determine new population sizes of wild type and mutant
  N_a_new <- max(offsp_a - mut_a_to_A, 0)
  N_A_new <- offsp_A + mut_a_to_A
  
  return(c(N_a_new, N_A_new))
}

# Initialize populations
N_a <- 100
N_A <- 0
sel_coeff <- 0
mut_rate <- 0

# Main simulation loop
for (i in 1:10) { # Run the simulation 10 times
  gen <- 0
  populations <- matrix(0, nrow = t_max, ncol = 2)
  populations[1, ] <- c(N_a, N_A)
  
  while (gen < t_max) {
    # Check if the current generation is within an event
    if (gen %% freq_event == 0 && (gen / freq_event) %% (dur_event + 1) < dur_event) {
      current_decay_rate <- decay_rate[1] # Decay rate during event
    } else {
      current_decay_rate <- decay_rate[2] # Decay rate without event
    }
    
    # Simulate population for the current generation
    populations[gen + 1, ] <- simulate_one_gen_2(populations[gen + 1, 1], populations[gen + 1, 2], current_decay_rate, 1.2, sel_coeff, mut_rate)
    
    # Increment generation
    gen <- gen + 1
  }
  
  # Store the result of this simulation
  simulation_results[[i]] <- populations
}

# Example of printing results from the first simulation
print(simulation_results[[1]])

```
```{r}
simulate_pop_2 <- function(N_init_a, N_init_A, decay_rate, sel_coeff, mut_rate, t_max) {
  # Create the vector in which to save the results
  pop_vector <- c(N_init_a, N_init_A)
  # initiate the variables
  pop_new <- c(N_init_a, N_init_A)
  
  # run the simulation until generation t_max
  for (i in 1:t_max+1) {
    # redefine the current population one generation later
    pop_new <- simulate_one_gen_2(pop_new[1],pop_new[2], decay_rate, sel_coeff, mut_rate)
    # add the new population sizes to the output vector
    pop_vector <- rbind(pop_vector,pop_new)
    # condition to stop the simulation before t_max: either the population exceeds 1.5 times the original population size, or it goes extinct
    if (pop_new[1]+pop_new[2]>=1.5*(N_init_a+ N_init_A) | pop_new[1]+pop_new[2]==0) break
  }
  
  # define the row and column names of the output vector
  rownames(pop_vector) <- (0:t_max)[1:length(pop_vector[,1])] # note that the vector has to be cut if the simulation stopped early
  colnames(pop_vector) <- c("a","A")
  # return the result
  return(pop_vector)	
}


# Test the function and plot the result
# set t_max to a value to be used througout
max_gen <- 1000
# create your simulation data, without mut_rate & sel_coeff and a higher decay_rate
output <- simulate_pop_2(1000,0,0.35,0.2,0,max_gen)
# show the last few lines of the data table
print(output)
# plot the output - note that the plot range and the x axis have to be adjusted based on the length of the table
# determine x axis range 
x_range <- 0:(length(output[,1])-1)
# this plots the total population size
plot(x_range,output[,1]+output[,2],type='l',ylim=c(0,max(output[,1]+output[,2])),xlab = "Generation",ylab = "Population size")
# add number of wild type individuals
lines(x_range,output[,1], col="blue")
# add number of mutant individuals
lines(x_range,output[,2], col="red")
```
```{r} 
# set some parameters to fixed values
init_a <- 100
init_A <- 0
m_rate <- 0
max_gen <- 1000
# determine how often to run the simulation for each set of parameters
no_replicates <- 100

# set parameters to vary
s_values <- c(0.2,0.3,0.4)
r_values <- c(0.35, 1.1)

# initialize data table - where to collect the results
data_table <- c()
# run the simulation across all chosen parameters
# loop over decay rates
for(rval in r_values){
  # loop over selection coefficients
  for(sval in s_values){
    # different way of running many simulations: make replicates using "repeat" function with a counter i
    # reset counter
    i<-1
    repeat {
      # increase counter by one
      i<-i+1
      # run the simulation once
      one_run <- simulate_pop_2(init_a,init_A,rval,sval,m_rate,max_gen)
      # determine total population sizes
      total_size <- one_run[,1]+one_run[,2]
      # determine minimum population size
      min_size <- min(total_size)
      # determine (first) generation at which this population size occurred
      min_gen <- as.numeric(which(total_size==min_size)[1])
      # enter the data into the table
      data_table <- rbind(data_table,c(rval,sval,min_gen,min_size)) # note that we add the varying parameters (decay rate and selection coefficient) to the table too
      # stop the repeated computation after no_replicates times
      if(i>no_replicates) break
    }
  }
}
# define column names of data table
colnames(data_table) <- c("r","s","min_gen","no_min")
# show the first lines of the output
print(head(data_table))
```

```{r}
repeat {
      # increase counter by one
      i<-i+1
      # run the simulation once
      run_one <- simulate_pop(init_a,init_A, decay_rate, sel_coeff, m_rate,t_max)
      # determine total population sizes
      total_size <- run_one[,1]+run_one[,2]
      # determine minimum population size
      min_size <- min(total_size)
      # determine (first) generation at which this population size occurred
      min_gen <- as.numeric(which(total_size==0))
      # enter the data into the table
      flooding_table <- rbind(flooding_table,c(min_gen,min_size)) # note that we add the varying parameters (decay rate and selection coefficient) to the table too
      # stop the repeated computation after no_replicates times
      if(i>replicates) break
    }
```

## Running replicates (works, but not clean)

```{r}

# set some parameters to fixed values
init_a <- 1000
init_A <- 0
m_rate <- 0.001
max_gen <- 1000
# determine how often to run the simulation for each set of parameters
no_replicates <- 100

# set parameters to vary
#s_values <- c(0.2,0.3,0.4)
#r_values <- c(0.05,0.1,0.15)

frequencies <- c(1:10)

# initialize data table - where to collect the results
data_table <- c()
#data_table_Nm <- c()
# run the simulation across all chosen parameters
# loop over decay rates
for(t_pause in frequencies){
  # loop over selection coefficients
  #for(sval in s_values){
    # different way of running many simulations: make replicates using "repeat" function with a counter i
    # reset counter
    i<-1
    repeat {
      # increase counter by one
      i<-i+1
      # run the simulation once
      one_run <- simulate_pop(init_a,init_A,0,0,m_rate,max_gen, t_pause)
      
      # determine total population sizes
      #total_size <- one_run[,1] + one_run [,2]
      
      # determine minimum population size
      #min_size <- min(total_size)
      # determine (first) generation at which this population size occurred
      #min_gen <- as.numeric(which(total_size==min_size)[1])
      
      final_size <- one_run[nrow(one_run),1] + one_run[nrow(one_run),2]
      final_gen <- length(one_run[,1])
      
      # enter the data into the table
      data_table <- rbind(data_table,c(t_pause,final_size,final_gen)) # note that we add the varying parameters (decay rate and selection coefficient) to the table too
      # stop the repeated computation after no_replicates times
      if(i>no_replicates) break
    }
  }
#}
# define column names of data table
colnames(data_table) <- c("f", "final_size","final_gen")
# show the first lines of the output
print(head(data_table))
```



## "Forever-Event" - Find the time to extinction

Simulating only the WT under "Flooding"-Event circumstances (higher decay rate). Without Mutant, to find the time it takes for the WT to go extinct.
This code is copied from the code block above (from Line 57).

```{r}
simulate_pop <- function(N_init_a, N_init_A, decay_rate, sel_coeff, mut_rate, t_max) {
  # Create the vector in which to save the results
  pop_vector <- c(N_init_a, N_init_A)
  # initiate the variables
  pop_new <- c(N_init_a, N_init_A)
  
  # run the simulation until generation t_max
  for (i in 1:t_max+1) {
    # redefine the current population one generation later
    pop_new <- simulate_one_gen(pop_new[1],pop_new[2], decay_rate, sel_coeff, mut_rate)
    # add the new population sizes to the output vector
    pop_vector <- rbind(pop_vector,pop_new)
    # condition to stop the simulation before t_max: either the population exceeds 1.5 times the original population size, or it goes extinct
    if (pop_new[1]+pop_new[2]>=1.5*(N_init_a+ N_init_A) | pop_new[1]+pop_new[2]==0) break
  }
```
  
## A small analysis

As you can see if you simulate and plot the output several times, there is a lot of stochasticity/variation in the outcome. Therefore, to get interpretable results, we have to run the simulation many times and summarize the outcome depending on the question of our study. Here I am showing an example in which we record the time and the value of the lowest population size (or the time of extinction, if the population size eventually goes to 0). Usually, we want to study this for a range of model parameters; here I am varying the decay rate and the selection coefficient. Other (maybe more interesting) features to study would be the following:

-   rescue probability dependent on initial frequency and selection coefficient
-   time at which the rescue mutation occurs first if it rescues, vs. if it doesn't rescue, dependent on initial population size and mutation rate
-   time until the population has recovered in case it becomes rescued, dependent on mutation rate and selection coefficient
-   ...

```{r}
#Original

# set some parameters to fixed values
init_a <- 100
init_A <- 0
m_rate <- 0.001
max_gen <- 1000
# determine how often to run the simulation for each set of parameters
no_replicates <- 100

# set parameters to vary
s_values <- c(0.2,0.3,0.4)
r_values <- c(0.05,0.1,0.15)

# initialize data table - where to collect the results
data_table <- c()
# run the simulation across all chosen parameters
# loop over decay rates
for(rval in r_values){
  # loop over selection coefficients
  for(sval in s_values){
    # different way of running many simulations: make replicates using "repeat" function with a counter i
    # reset counter
    i<-1
    repeat {
      # increase counter by one
      i<-i+1
      # run the simulation once
      one_run <- simulate_pop(init_a,init_A,rval,sval,m_rate,max_gen)
      # determine total population sizes
      total_size <- one_run[,1]+one_run[,2]
      # determine minimum population size
      min_size <- min(total_size)
      # determine (first) generation at which this population size occurred
      min_gen <- as.numeric(which(total_size==min_size)[1])
      # enter the data into the table
      data_table <- rbind(data_table,c(rval,sval,min_gen,min_size)) # note that we add the varying parameters (decay rate and selection coefficient) to the table too
      # stop the repeated computation after no_replicates times
      if(i>no_replicates) break
    }
  }
}
# define column names of data table
colnames(data_table) <- c("r","s","min_gen","no_min")
# show the first lines of the output
print(head(data_table))
```

## Replicating "Forever-Event" for several generations

Finding the average time for the WT to go extinct in an event.
The code simulates several generations of events and returns the average of how fast the WT population goes extinct, when we have a higher decay rate (r_values in line 225)
This code is copied from the code block above (from Line 159).

```{r}
# set some parameters to fixed values, m_rate is 0
init_a <- 100
init_A <- 0
m_rate <- 0
max_gen <- 1000
# determine how often to run the simulation for each set of parameters
no_replicates <- 750

# set parameters to vary, r_values (decay) is fixed
s_values <- c(0.2,0.3,0.4)
r_values <- 0.01

# initialize data table - where to collect the results
data_table <- c()
# run the simulation across all chosen parameters
# loop over decay rates
for(rval in r_values){
  # loop over selection coefficients
  for(sval in s_values){
    # different way of running many simulations: make replicates using "repeat" function with a counter i
    # reset counter
    i<-1
    repeat {
      # increase counter by one
      i<-i+1
      # run the simulation once
      one_run <- simulate_pop(init_a,init_A,rval,sval,m_rate,max_gen)
      # determine total population sizes
      total_size <- one_run[,1]+one_run[,2]
      # determine minimum population size
      min_size <- min(total_size)
      # determine (first) generation at which this population size occurred
      min_gen <- as.numeric(which(total_size==0))
      # enter the data into the table
      data_table <- rbind(data_table,c(rval,sval,min_gen,min_size)) # note that we add the varying parameters (decay rate and selection coefficient) to the table too
      # stop the repeated computation after no_replicates times
      if(i>no_replicates) break
    }
  }
}
# define column names of data table
colnames(data_table) <- c("r","s","min_gen","no_min")
# show the first lines of the output
print(head(data_table))
#mean of generations after which the WT population goes extinct
#mean.extWT <- mean(min_gen)

#print(mean.extWT)
```

## Implement several Events

The goal: Find a range of interesting frequencies of events at which the WT population starts struggling or goes extinct.
This code simulate a generation after another, depending on in which generation it starts. A list contains several starts of Events which then determines the decay rate at which the generation is simulated. This code only works for one generation and ignores, that an events lasts longer than one generation.
The code was written from scratch and copied from the first code block (about the simulate_one_gen)

```{r}
decay_rate = c(0.35, 1.1) #decay rate in presence and absence of event --> strength of event
freq_event = 10 #every 10 generations an event happens
dur_event = 2 #event has a duration of 2 generations
t_max = 100
gen = 0
start_event = list()

while (gen < t_max){
  start_event <- append(start_event, gen)
  gen = gen + dur_event + freq_event
}
#print(start_event)
#start_event[1]
#start_event[2]
#start_event[3]
#start_event[4]
#start_event[5]
#start_event <- c(start_event,start_event+1,start_event+2)
simulate_one_gen_2 <- function(N_a, N_A, decay_rate, growth_rate, sel_coeff, mut_rate, gen) {
  if(gen %in% start_event) {
    decay_rate = 0.35
    growth_rate = 0
  }
  else{
      decay_rate = 0
      growth_rate = 1.1
  }
  offsp_a <- rpois(1, N_a * (1-decay_rate))
  offsp_A <- rpois(1, N_A * (1-decay_rate+sel_coeff))
  # draw new mutants according to Poisson distribution
  mut_a_to_A <- rpois(1, offsp_a * mut_rate)
  
  # determine new population sizes of wild type and mutant
  N_a_new <- max(offsp_a - mut_a_to_A, 0)
  N_A_new <-  offsp_A + mut_a_to_A
  
  return(c(N_a_new, N_A_new))
}

print(simulate_one_gen_2(100, 0, 0, 1.2, 0, 0, 3))


#This was just to test, did not work
#for (i in 1:t_max+1) {
 # new_pop <- simulate_one_gen_2(N_a, N_A, decay_rate, growth_rate, sel_coeff, mut_rate, gen)
  #vector_pop <- rbind(vector_pop, new_pop)
  #if(new)


```


```{r}
  # define the row and column names of the output vector
  rownames(pop_vector) <- (0:t_max)[1:length(pop_vector[,1])] # note that the vector has to be cut if the simulation stopped early
  colnames(pop_vector) <- c("a","A")
  # return the result
  return(pop_vector)	
}


# Test the function and plot the result
# set t_max to a value to be used througout
max_gen <- 1000
# create your simulation data
output <- simulate_pop(1000,0,0.35,0,0,max_gen) #mut rate and initial mutation pop are 0 to time the decay of WT population
# show the last few lines of the data table
print(output)
# plot the output - note that the plot range and the x axis have to be adjusted based on the length of the table
# determine x axis range 
x_range <- 0:(length(output[,1])-1)
# this plots the total population size
plot(x_range,output[,1]+output[,2],type='l',ylim=c(0,max(output[,1]+output[,2])),xlab = "Generation",ylab = "Population size")
# add number of wild type individuals
lines(x_range,output[,1], col="blue")
# add number of mutant individuals
lines(x_range,output[,2], col="red")
```


```{r}
# Set parameters to fixed values
init_a <- 1000
init_A <- 0
m_rate <- 0.001
max_gen <- 1000
no_replicates <- 100

# Frequencies to vary
frequencies <- c(10:20)

# Initialize data table
data_table <- matrix(nrow = 0, ncol = 3) # For 3 columns: t_pause, final size, final gen

# Run the simulation across all chosen parameters
for (t_pause in frequencies) {
  for (replicate in 1:no_replicates) {
    # Run the simulation once
    one_run <- simulate_pop(init_a, init_A, 0.03, 0.05, m_rate, max_gen, t_pause)
    
    # Determine total population sizes
    final_size <- one_run[nrow(one_run), 1] + one_run[nrow(one_run), 2]
    final_gen <- nrow(one_run) - 1 # Subtracting 1 because rownames start from 0
    
    # Enter the data into the table
    data_table <- rbind(data_table, c(t_pause, final_size, final_gen))
  }
}

# Define column names of data table
colnames(data_table) <- c("t_pause", "final_size", "final_gen")

# Convert data_table to a data frame for better handling
data_table <- as.data.frame(data_table)
data_table$t_pause <- as.numeric(as.character(data_table$t_pause))
data_table$final_size <- as.numeric(as.character(data_table$final_size))
data_table$final_gen <- as.numeric(as.character(data_table$final_gen))

# Show the first lines of the output
print(head(data_table))

```


## After discussion with Loic - see photo on repository

The timeline is separated into phases, which indicate where there is an event or not. for example; odd phase-number indicate no event and even phase-number indicates an event. The following code should simulate the population according to the phase and check if the phase number corresponds to an event or not.
The code was produced after a discussion with Loic.
```{r}
t_pause = 10 #t_switch
t_event = 2
t_switch = t_pause
ph = 1
sel_coeff = 0.0

replicates <- 100

flooding_table <- c()


for (i in 1:t_max+1){
  if(ph %%2 == 0){
    decay_rate = 0.3
  }else{
    decay_rate = 0
  }

  if (i >= t_switch) {
    if(ph %% 2 == 0) {
      t_switch = t_switch + t_event
      ph = ph + 1
    }else{
      t_switch = t_switch + t_pause
      ph = ph + 1
    }
  }
}

colnames(flooding_table) <- c("N_a_new", "N_A_new")

#print(simulate_one_gen_3(100, 0, 0, 1.2, 0, 0, 3))
print(head(flooding_table))


```

```{r}

# set some parameters to fixed values
init_a <- 1000
init_A <- 0
m_rate <- 0.001
max_gen <- 1000
# determine how often to run the simulation for each set of parameters
no_replicates <- 100

# set parameters to vary
frequencies <- c(1:10)

# initialize data table - where to collect the results
data_table_Pop <- c()
data_table_Nw <- c()
data_table_Nm <- c()

# run the simulation across all chosen parameters
# loop over different frequencies
for(t_pause in frequencies){

    # different way of running many simulations: make replicates using "repeat" function with a counter i
    # reset counter
    i<-1
    repeat {
      # increase counter by one
      i<-i+1
      # run the simulation once
      one_run <- simulate_pop(init_a,init_A,0,0,m_rate,max_gen, t_pause)
      
      #determine the final size of the population and number of generations it took for the simulation to stop
      final_size <- one_run[nrow(one_run),1] + one_run[nrow(one_run),2]
      final_gen <- length(one_run[,1])
      
      # enter the data into the table
      data_table_Pop <- rbind(data_table_Pop,c(t_pause,final_size,final_gen)) 
      
      size_wt <- one_run[nrow(one_run),1]
      size_mut <- one_run[nrow(one_run),2]
      
         # Count surviving populations
    if (size_wt > 0) {
      surviving_Nw[t_pause] <- surviving_Nw[t_pause] + 1
    }
    if (size_mut > 0) {
      surviving_Nm[t_pause] <- surviving_Nm[t_pause] + 1
    }
      
      data_table_Nw <- rbind(data_table_Nw,c(t_pause,size_wt,final_gen))
      data_table_Nm <- rbind(data_table_Nm,c(t_pause,size_mut,final_gen))
      
      # stop the repeated computation after no_replicates times
      if(i>no_replicates) break
    }
  }
#}
# define column names of data table
colnames(data_table_Pop) <- c("f", "Final Pop","final_gen")
colnames(data_table_Nw) <- c("f", "Final WT","final_gen")
colnames(data_table_Nm) <- c("f", "Final Mut","final_gen")
# show the first lines of the output
print(head(data_table_Pop))
```

